Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-01 rnni_convexity.tex
pass to cluster thm

With the results of Section~\ref{section:caterpillar_convex} we can easily get the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in Theorem 7 of \autocite{Gavryushkin2018-ol}.
In the following we show that the upper bound of the diameter provided by Lemma~\ref{lemma:diameter_bound} actually is the exact diameter of $\rnni$.

@@

\begin{proof}
Let $T = [1,2,3,\ldots,n]$ and $R = [n-1,n,n-2,n-3, \ldots, 1]$ be two caterpillar trees.
According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i = 1, \ldots, n-1$) swaps $n-1-i$ times with its upper neighbour.
Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i)  = \frac{(n-1)(n-2)}{2}$.
\end{proof}

@@

Induction basis: $n = 3$\\
$\mdtree(T)$ results either in $R =[1,3,2]$ or $R'= [2,3,1]$ for $T = [1,2,3]$.
Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.

Induction step: $n \to n+1$\\
Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa $\mdtree$ computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
Let us now assume that $T$ is a tree on $n+1$ taxa.
In the list $L$ computed in $\mdtree$ (Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alg:max_dist_tree}) taxon $n+1 $ is the last one as its parent, the root, has the highest rank in $T$.
This means that this taxon $n+1$ is the last one added to $R$ as in Line~\ref{alg:mdtree_lineAddTaxon} of this algorithm and therefore, it has parent with rank one in $R$.
Because the parent of $n+1$ has rank $n$ in $T$ and rank $1$ in $R$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
$R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed in Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alg:max_dist_tree} for $T|_n$ is the same as the one for $T$, but without the last element.
So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
\end{proof}

In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
In fact, the output tree of $\mdtree$ can have any tree topology, since there is no restriction on where to add the new taxon in the already existing tree (Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{alg:max_dist_tree}).
This means that for each caterpillar input tree $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
And as permuting labels of trees does not change the distance between them, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
By this we obtain Corollary~\ref{corollary:radius}.

>>>>

In this section we establish diameter and radius of the $\rnni$ graph.
In Theorem 7 of \autocite{Gavryushkin2018-ol} an upper bound for the diameter is given, but we are able to give the exact diameter using the results of Section~\ref{section:caterpillar_convex}.
In the following we prove the exact diameter of $\rnni$ before we investigate the radius of $\rnni$.

@@

\begin{proof}
Let $T = [1,2,3,\ldots,n]$ and $R = [n-1,n,n-2,n-3, \ldots, 1]$ be two caterpillar trees.
According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i = n, \ldots, 3$) swaps $i-2$ times with its upper neighbour.
Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=3}^{n}(i-2)  = \frac{(n-1)(n-2)}{2}$.
\end{proof}

@@

The lemma is true for the induction basis on $n=3$ taxa since all three trees have distance one, which equals diameter and radius in this case.

Let us now assume for the induction step that $T$ is a tree on $n+1$ taxa.
Taxon $n+1$ is the last one in the list $L$ computed in $\mdtree$ as its parent in $T$ is the root.
It follows that this taxon $n+1$ is the last one added to $R$ in the algorithm, so its parent has rank one in $R$.
With Lemma~\ref{lemma:distance_delete_taxon} we can follow that for $T$ and $R$ restricted to taxa $\{1, \ldots, n\}$ it holds $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
$R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed for $T|_n$ in $\mdtree$ is the same as the one for $T$, but without the last element.
So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
\end{proof}

In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
In fact, the output tree of $\mdtree$ can have any tree topology, since in each step of the algorithm the next taxon can be added on any edge incident to a leaf in the tree.
Therefore, if the input tree is a caterpillar tree, $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
And as permuting labels of trees does not change the distance between them, there exists for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
By this we obtain Corollary~\ref{corollary:radius}.
